Supplementary material to the paper 'A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast

### 2022-09-14 analysis/paper/paper.Rmd
Adding a couple more sites for re-dating and updating supplementary material references

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed shoreline dating which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous hypotheses that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on identifying a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). This area has newly compiled displacement curves for Horten [@romundset2021], Larvik [@sørensen2014; @sørensen2014b; @sorensen2023], Tvedestrand [@romundset2018; @romundset2018b], and Arendal [@romundset2018b].

@@

```{r map-iso,  fig.cap = paste("A) Location of the study area and the distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\\circ}$ (Romundset et al. 2018, although see Sørensen et al. 2014), B) Displacement curves. Note the increasing steepness of the curves towards the north-east."), fig.height = 13, fig.width = 12, regfloat=TRUE}
# Create a bounding box around site points to limit the overview map
sitbbox <- st_bbox(sites_sa)
sitbbox[1:2] <- sitbbox[1:2] - 300000
sitbbox[3:4] <- sitbbox[3:4] + 300000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

# Reproject the bounding box to match world map, and crop the world map
# with the bounding box polygon
bound_reproj <- st_transform(boundingpoly, st_crs(bmap))
mapcountries <- bmap %>% 
  filter(st_intersects(., bound_reproj, sparse = FALSE))
count_reproj <- st_transform(mapcountries, st_crs(sites_sa))

@@

cpts <- filter(cpts, !(NAME %in% c("Tvedestrand", "Porsgrunn")))

@@

disp_curves <- ggplot(displacement_curves, aes(x = years, col = name)) +
  geom_line(aes(y =  upperelev)) +
  geom_line(aes(y = lowerelev)) +
  labs(title = "B", y = "Meters above present sea-level", x = "BCE/CE") +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange")) +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed using the Sum function. Multiple phases at a single site were treated as independent of each other.

The excavation of archaeological sites in Norway typically occur in advance of residential and commercial infrastructure development. As the data collection for the utilised DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation. In addition to employing the 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension [e.g. @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021]. In addition to code and original spatial data being available in the online repository for the paper, the analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.

@@

curve_plot <- ggplot2::ggplot() + 
              geom_line(data = displacement_curves, 
                      aes(x = years, y = upperelev, col = name), alpha = 0.4) +
  geom_line(data =  displacement_curves, 
                      aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
  geom_line(data = sitecurve,
            aes(x = years, y = upperelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = lowerelev, col = name)) +
  scale_colour_manual(values = c("Arendal" = "black",
                                   "Larvik" = "darkgreen",
                                   "Tvedestrand" = "blue",
                                   "Horten" = "darkorange",
                                 "Hegna vest 1" = "red")) +
  ylab("Meters above present sea-level") +
  xlab("BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

We can then get rid of parameter $D$ by summaing all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:

@@

To evaluate the outlined procedure it is used to shoreline date the sites from where the method was derived to check if the resulting shoreline dates correspond to the radiocarbon dates of the sites (Figure \@ref(fig:shoredate)). The Late Neolithic sites are also included here for illustrative purposes, even though these have not informed the decay ratio in use. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits. The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008. Here, 100,000 age samples drawn from the probability distribution of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)). When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date corresponds to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%). Only including dates modelled to be older than 2500 BCE with 95% probability, i.e. older than the Late Neolithic, improves this to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%). When only including dates older than 4000 BCE with 95% probability, i.e. only Mesolithic site phases, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%). The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).

@@

Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent in this reconstruction.

>>>>

---
title: "A simulation-based assessment of the relation between Stone Age sites and relative sea-level change along the Norwegian Skagerrak coast"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author: 
- 'Isak Roalkvam'
- 'University of Oslo, Institute of Archaeology, Conservation and History'
  # - Isak Roalkvam:
  #     email: isak.roalkvam@iakh.uio.no
  #     institute: [UiO]
  #     correspondence: true
# institute:
#   - UiO: University of Oslo, Institute of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      # - --lua-filter=../templates/scholarly-metadata.lua
      # - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  A central premise for the Stone Age archaeology of northern Scandinavia is that most coastal sites were located on or close to the contemporary shoreline when they were in use. By reconstructing the trajectory of rapid and continuous relative sea-level fall that characterises large regions of Fennoscandia, this offers a dating method termed 'shoreline dating' which is widely applied. However, while the potentially immense benefits of an additional source of temporal data separate from radiometric and typological methods is unquestionable, the geographical contingency and thus relative rarity of the method means that it has been under limited scrutiny compared to more established dating techniques in archaeology. This paper attempts to remedy this by quantifying the spatial relationship between Stone Age sites located beneath the marine limit and the prehistoric shoreline along the Norwegian Skagerrak coast. Monte Carlo simulation is employed to combine the uncertainty associated with independent temporal data on the use of the sites in the form of ^14^C-dates and the reconstruction of local shoreline displacement. The findings largely confirm previous hypotheses that sites older than the Late Neolithic tend to have been located on or close to the shoreline when they were occupied. Drawing on the quantitative nature of the results, a new and formalised method for the shoreline dating of sites in the region is proposed and compared to previous applications of the technique.

@@

# Data
To get at the relationship between sites and the contemporaneous shoreline, this analysis was dependent on identifying a study area with good control of the trajectory of prehistoric shoreline displacement. While there is displacement data available for other areas of south-eastern Norway [e.g. @hafsten1957; @sørensen1979; @sørensen1999], considerable methodological developments in recent years means that the most well-established displacement curves are from the region stretching from Horten county in the north-east, to Arendal in the south-west (Figure \@ref(fig:map-iso)). This area has newly compiled displacement curves for Skoppum in Horten [@romundset2021], Gunnarsrød in Porsgrunn [@sørensen2014; @sørensen2014b; @sorensen2023], Hanto in Tvedestrand [@romundset2018; @romundset2018b], and Bjørnebu in Arendal [@romundset2018b].

@@

```{r map-iso,  fig.cap = paste("A) Location of the study area and the distribution of the ", nrow(sites_sa), " analysed sites relative to the isobases of the displacement curves. The isobases have a direction of 327$^{\\circ}$ (Romundset et al. 2018, although see Sørensen et al. 2014), B) Displacement curves. Note the increasing steepness of the curves towards the north-east."), fig.height = 13, fig.width = 12, regfloat=TRUE}
# To make sure labelling ends up the same every knit.
set.seed(1)

# Create a bounding box around site points to limit the overview map.
sitbbox <- st_bbox(sites_sa)
sitbbox[1:2] <- sitbbox[1:2] - 100000
sitbbox[3:4] <- sitbbox[3:4] + 100000
boundingpoly <- st_as_sf(st_as_sfc(sitbbox))

# Reproject the bounding box to match world map, and crop the world map
# with the bounding box polygon.
bound_reproj <- st_transform(boundingpoly, st_crs(bmap))
mapcountries <- bmap %>% 
  filter(st_intersects(., bound_reproj, sparse = FALSE))
count_reproj <- st_transform(mapcountries, st_crs(sites_sa))

@@

@@

# Rename to use the precise placenames
rname_disp <- displacement_curves
rname_disp$name <- ifelse(displacement_curves$name == "Larvik",
                                   "Gunnarsrød", 
                                   ifelse(
                                     displacement_curves$name == "Tvedestrand",
                                     "Hanto", 
                                     ifelse(
                                       displacement_curves$name == "Arendal",
                                       "Bjørnebu", "Skoppum")))

disp_curves <- ggplot(rname_disp, aes(x = years, col = name)) +
  geom_line(aes(y =  upperelev)) +
  geom_line(aes(y = lowerelev)) +
  labs(title = "B", y = "Meters above present sea-level", x = "BCE/CE") +
  scale_colour_manual(values = c("Skoppum" = "darkorange",
                                 "Gunnarsrød" = "darkgreen",
                                 "Hanto" = "blue",
                                 "Bjørnebu" = "black")) +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

Shoreline dating is based on the spatial relationship between two phenomena, occupation of sites and shoreline displacement, each associated with temporal uncertainty. The first task was therefore to ascribe a likely date and associated degree of uncertainty to these dimensions. To take account of the gradient in the isostatic rebound, the trajectory of shoreline displacement was first interpolated to each site location based on the distance to the isobases of the displacement curves, using inverse distance weighting [e.g. @conolly2020; @conolly2006, 94--97]. This was done for each year along the entirety of the curves, weighting the interpolation by the squared inverse of the distances. The result of this process is shown for an example site in Figure \@ref(fig:example-1). For the sites all radiocarbon dates were first individually calibrated using the IntCal20 calibration curve [@reimer2020] using OxCal v4.4.4 [@bronkramsey2009] through the oxcAAR package for R [@hinz2021]. Radiocarbon dates associated with each site were then grouped if their date ranges intersected at 99.7% probability, meaning these were effectively taken to be associated with the same event, here termed settlement or site phase. In the case where there are multiple dates believed to belong to a single settlement phase, these were modelled using the Boundary function in OxCal and then summed using the Sum function. Multiple phases at a single site were treated as independent of each other.

The excavation of archaeological sites in Norway typically occur in advance of residential and commercial infrastructure development. As the data collection for the utilised DTM was begun by the Norwegian Mapping Authority in 2016, the area of the DTM immediately surrounding the sites has sometimes been severely impacted by disturbances after the excavation. In addition to employing the 10m resolution DTM to alleviate some of these issues, this also necessitated some additional editing of the elevation raster. This involved manually defining the extent of problem areas such as railways, highways, quarries and the like. The DTM values on these were then set to missing, and new elevation values were interpolated from the surrounding terrain. This was done using regularised spline interpolation with tension [e.g. @conolly2020], using the default settings of r.fillnulls from GRASS GIS [@grass2017] in R through the package rgrass7 [@bivand2021]. In addition to code and original spatial data being available in the online repository for the paper, the location and analysis of each individual site is presented in the supplementary material where it has been noted when the area surrounding a site has been edited in this manner.

@@

curve_plot <- ggplot2::ggplot() + 
              geom_line(data = rname_disp, 
                      aes(x = years, y = upperelev, col = name), alpha = 0.4) +
  geom_line(data =  rname_disp, 
                      aes(x = years, y = lowerelev, col = name), alpha = 0.4) +
  geom_line(data = sitecurve,
            aes(x = years, y = upperelev, col = name)) +
  geom_line(data = sitecurve,
            aes(x = years, y = lowerelev, col = name)) +
  scale_colour_manual(values = c("Skoppum" = "darkorange",
                                 "Gunnarsrød" = "darkgreen",
                                 "Hanto" = "blue",
                                 "Bjørnebu" = "black",
                                 "Hegna vest 1" = "red")) +
  ylab("Meters above present sea-level") +
  xlab("BCE/CE") +
  theme_bw() +
  theme(legend.title = element_blank(), legend.position = "bottom",
        legend.direction = "horizontal")

@@

We can then get rid of parameter $D$ by summing all possible distances between site and the shoreline. Given its elevation, the probability for the date of the occupation of a site is then:

@@

To evaluate the outlined procedure it is used to shoreline date the sites from where the method was derived to check if the resulting shoreline dates correspond to the radiocarbon dates associated with the sites (Figure \@ref(fig:shoredate)). The Late Neolithic sites are also included here for illustrative purposes, even though these have not informed the decay ratio in use. Following from having defined the distance between intersecting sea- and site polygons as zero during simulations, the sites were dated using the mean elevation of the site polygons to allow for some variation in elevation over the site limits. The synchroneity between radiocarbon and shoreline dates was then evaluated using the method presented by @parnell2008. Here, 100,000 age samples drawn from the probability distribution of each shoreline date were subtracted from 100,000 age samples drawn from the corresponding modelled ^14^C-dates. The resulting range of the 95% highest density region [HDR, @hyndman1996] was then checked to see if it crosses zero, in which case the dates are considered to be in agreement (Figure \@ref(fig:shoredatediff)). When excluding the earliest occupation phase at Gunnarsrød 5, the deviation of which is to be expected based on issues with the DTM (see above), the shoreline date corresponds to the radiocarbon dates in `r synchdata[synchdata$cross == "true", "row_total"]` out of `r sum(synchdata$row_total)-1` cases (`r round(synchdata[synchdata$cross == "true", "row_total"]/sum(synchdata$row_total)*100)`%). Only including dates modelled to be older than 2500 BCE with 95% probability, i.e. older than the Late Neolithic, improves this to `r cormeso + corpreln` out of `r phasesums["meso"] + phasesums["pre_ln"]` cases (`r round((cormeso + corpreln)  / (phasesums["meso"] + phasesums["pre_ln"])*100)`%). When only including dates older than 4000 BCE with 95% probability, i.e. only Mesolithic site phases, the success rate is further increased to `r cormeso`/`r phasesums["meso"]` (`r round(cormeso/phasesums["meso"]*100)`%). The three failed Mesolithic shoreline dates are from the early sites Langemyr and Kvastad A2, with the likely implication that a lower decay ratio than what is used for characterising the distance between site and shoreline for all sites in aggregate should be used for sites known to be from the earliest part of the Mesolithic (see also Figure \@ref(fig:results)).

@@

Finally, this analysis employed a simulation approach to integrate multiple sources of spatio-temporal uncertainty. Here this was simply used to inform the question of the distance between sites and the shoreline. However, this method and general framework can be extended to a wide range of use-cases where one needs to visualise, and quantitatively or qualitatively evaluate the relationship between archaeological phenomena, the prehistoric shoreline, and the uncertainty inherent to this reconstruction.

### 2022-09-14 analysis/supplementary-materials/supplementary.Rmd
Adding a couple more sites for re-dating and updating supplementary material references

The artefact inventory from Løvås 1 matches the radiocarbon dates [@reitan2022]. No editing of the raster appeared to be necessary. Interestingly only the Løvås sites were simulated to be situated beneath the sea-level. The typologically secure inventory combined with the many overlapping dates from various features does indicate that the issue is with the displacement curve and not the dating of the site. Similar situation to that on Løvås 2 and 3 (below).

@@

The ^14^C-dates matches the lithic inventory from Stokke/Polland 5 [@mansrud2017]. The site is located where the highway runs today. This appears to have been successfully edited. The excavated area is more extensive (field A--C in the report) that what is used as the site limit here, but the relation between the different areas is not clear. Field A, used as the site limit here, holds all the dated features.

@@

Typology of lithics match the ^14^C-dates from Torstvet [@mansrud2013b]. The site is located where the highway runs today. This appears to have been successfully edited, although there is a possibility that this has led to the sea-level being simulated sightly further away from the site than what would otherwise have been the case. However, as with he Hovland-sites above, this appears to rather follow from the general topography of the landscape which makes the reconstruction more uncertain than in areas with a higher relief.

@@

Vallermyrene 1 was divided into 1a and 1b based on elevation difference, typology and 14C-dates [@reitan2014f]. Vallermyene 1a is the oldest of the two. Although situated not that far from the railway, the area immediatly surrounding the site has not been disturbed in a way that appears to impact the reconstruction of the past sea-level.

@@

Vallermyrene 2 was a large site situated in a field [@reitan2014o]. The site has ^14^C-dates and artefacts dating occupations at the sites to multiple phases throughout prehistory, including the Bronze Age and phases up until the Viking Age. The Stone Age activity on the site was evidenced by 359 lithics and a fireplace dated to the Early Neolithic. Given it's uncertain extent, the site limit used here is defined by the fireplace.

@@

<!-- @nyland2012b -->
<!-- @danielsen2018 -->
<!-- @koxvold2018 -->
<!-- @solheim2017 -->
<!-- @granum2018 -->
<!-- @reitan2014l -->
<!-- @reitan2014f -->
<!-- @carrasco2014 -->
<!-- @carrasco2014 -->
<!-- @fossum2014b -->
<!-- @fossum2014c -->
<!-- @koxvold2017l -->
<!-- @koxvold2017n -->
<!-- @havstein2017a -->
<!-- @havstein2017 -->
<!-- @koxvold2017h -->
<!-- @eigeland2017a -->
<!-- @koxvold2013 -->
<!-- @koxvold2017d -->
<!-- @koxvold2017e -->
<!-- @koxvold2016 -->
<!-- @koxvold2017m -->
<!-- @koxvold2017f -->
<!-- @koxvold2017g -->
<!-- @koxvold2020a -->
<!-- @stokke2018a -->
<!-- @reitan2018a -->
<!-- @stokke2018a -->
<!-- @nielsen2017 -->
<!-- @nielsen2017 -->
<!-- @darmark2018c -->
<!-- @viken2018a -->
<!-- @darmark2018b -->
<!-- @eymundsson2014 -->
<!-- @eggen2014b -->
<!-- @reitan2014h -->
<!-- @reitan20140 -->
<!-- @mansrud2018 -->
<!-- @persson2008 -->
<!-- @eigeland2014 -->
<!-- @eigeland2014a -->
<!-- @fossum2014a -->
<!-- @olsen2013a -->
<!-- @koxvold2013a -->
<!-- @mansrud2013c -->
<!-- @hårstad2021 -->
<!-- @berge2012 -->
<!-- @schallerahrberg2012 -->
<!-- @nyland2012 -->
<!-- @amundsen2012 -->
<!-- @nyland2012a -->
<!-- @amundsen2012a -->
<!-- @jaksland2012 -->
<!-- @jaksland2012a -->
<!-- @carrasco2015 -->
<!-- @carrasco2015 -->
<!-- @persson2014a -->
<!-- @hårstad2021b -->
<!-- @koxvold2020 -->
<!-- @viken2018 -->
<!-- @darmark2018a -->
<!-- @darmark2018 -->
<!-- @darmark2018 -->
<!-- @darmark2018d -->
<!-- @nielsen2017a -->
<!-- @ekstrand2013 -->
<!-- @ekstrand2013 -->
<!-- @ekstrand2013 -->
<!-- @amundsen2012b -->
<!-- @fossum2014 -->
<!-- @fossum2017c -->
<!-- @mansrud2016 -->
<!-- @koxvold2017j -->
<!-- @fossum2017d -->
<!-- @eggen2014 -->
<!-- @melvold2014a -->
<!-- @stokke2017 -->
<!-- @stokke2017 -->
<!-- @koxvold2017c -->
<!-- @koxvold2017b -->
<!-- @koxvold2017a -->
<!-- @reitan2021 -->
<!-- @reitan2021 -->
<!-- @reitan2014i -->
<!-- @reitan2014f -->

>>>>

The artefact inventory from Løvås 1 matches the radiocarbon dates [@reitan2022]. No editing of the raster appeared to be necessary. Interestingly only the Løvås sites were simulated to be situated beneath the sea-level. The typologically secure inventory combined with the many overlapping dates from various features does indicate that the issue is with the displacement curve and not the dating of the site (see also main text). Similar situation to that on Løvås 2 and 3 (below).

@@

The ^14^C-dates matches the lithic inventory from Stokke/Polland 5 [@mansrud2017]. The site is located where the highway runs today. This appears to have been successfully edited. The excavated area is more extensive (field A--C in the report) than what is used as the site limit here, but the relation between the different areas is not clear. Field A, used as the site limit here, holds all the dated features.

@@

Typology of lithics match the ^14^C-dates from Torstvet [@mansrud2013b]. The site is located where the highway runs today. This appears to have been successfully edited, although there is a possibility that this has led to the sea-level being simulated sightly further away from the site than what would otherwise have been the case. However, as with he Hovland sites above, this appears to rather follow from the general topography of the landscape characterised by a low relief which makes the reconstruction more uncertain.

@@

Vallermyrene 1 was divided into 1a and 1b based on elevation difference, typology and ^14^C-dates [@reitan2014f]. Vallermyrene 1a is the oldest of the two. Although situated not that far from the railway, the area immediately surrounding the site has not been disturbed in a way that appears to impact the reconstruction of the past sea-level.

@@

Vallermyrene 2 was a large site situated in a field [@reitan2014o]. The site has ^14^C-dates and artefacts dating occupations at the site to multiple phases throughout prehistory, including the Bronze Age and phases up until the Viking Age. The Stone Age activity on the site was evidenced by 359 lithics and a fireplace dated to the Early Neolithic. Given it's uncertain extent, the site limit used here is defined by the fireplace.

@@

<!-- @nyland2012b -->
<!-- @danielsen2018 -->
<!-- @koxvold2018 -->
<!-- @solheim2017 -->
<!-- @granum2018 -->
<!-- @reitan2014l -->
<!-- @reitan2014f -->
<!-- @carrasco2014 -->
<!-- @carrasco2014 -->
<!-- @fossum2014b -->
<!-- @fossum2014c -->
<!-- @koxvold2017l -->
<!-- @koxvold2017n -->
<!-- @havstein2017a -->
<!-- @havstein2017 -->
<!-- @koxvold2017h -->
<!-- @eigeland2017a -->
<!-- @koxvold2013 -->
<!-- @koxvold2017d -->
<!-- @koxvold2017e -->
<!-- @koxvold2016 -->
<!-- @koxvold2017m -->
<!-- @koxvold2017f -->
<!-- @koxvold2017g -->
<!-- @koxvold2020a -->
<!-- @stokke2018a -->
<!-- @reitan2018a -->
<!-- @stokke2018a -->
<!-- @nielsen2017 -->
<!-- @nielsen2017 -->
<!-- @darmark2018c -->
<!-- @viken2018a -->
<!-- @darmark2018b -->
<!-- @eymundsson2014 -->
<!-- @eggen2014b -->
<!-- @reitan2014h -->
<!-- @reitan20140 -->
<!-- @mansrud2018 -->
<!-- @persson2008 -->
<!-- @eigeland2014 -->
<!-- @eigeland2014a -->
<!-- @fossum2014a -->
<!-- @olsen2013a -->
<!-- @koxvold2013a -->
<!-- @mansrud2013c -->
<!-- @hårstad2021 -->
<!-- @berge2012 -->
<!-- @schallerahrberg2012 -->
<!-- @nyland2012 -->
<!-- @amundsen2012 -->
<!-- @nyland2012a -->
<!-- @amundsen2012a -->
<!-- @jaksland2012 -->
<!-- @jaksland2012a -->
<!-- @carrasco2015 -->
<!-- @carrasco2015 -->
<!-- @persson2014a -->
<!-- @hårstad2021b -->
<!-- @koxvold2020 -->
<!-- @viken2018 -->
<!-- @darmark2018a -->
<!-- @darmark2018 -->
<!-- @darmark2018 -->
<!-- @darmark2018d -->
<!-- @nielsen2017a -->
<!-- @ekstrand2013 -->
<!-- @ekstrand2013 -->
<!-- @ekstrand2013 -->
<!-- @amundsen2012b -->
<!-- @fossum2014 -->
<!-- @fossum2017c -->
<!-- @fossum2017b -->
<!-- @mansrud2016 -->
<!-- @koxvold2017j -->
<!-- @fossum2017d -->
<!-- @eggen2014 -->
<!-- @melvold2014a -->
<!-- @stokke2017 -->
<!-- @stokke2017 -->
<!-- @koxvold2017c -->
<!-- @koxvold2017b -->
<!-- @koxvold2017a -->
<!-- @reitan2021 -->
<!-- @reitan2021 -->
<!-- @reitan2014i -->
<!-- @reitan2014f -->
<!-- @reitan2014s -->
